$p$-adic precision

### 2014-02-05 padicprec.tex
Sections 4 and 5.2 moved to appendices. The paper (without appendices) now fits in 15 pages as wanted by ANTS.

Lemma \ref{lem:main} and Proposition \ref{prop:locanalytic} have important 
applications to effective computations with $p$-adic numbers or power 
series, that we are going to discuss now. Let us recall first that a 
datum involving elements of a complete ultrametric field $K$ (\emph{e.g.} 
$p$-adic numbers or power series) cannot be stored entirely on a computed 
because they basically convey an infinite amount of informations. That is 
why, instead of this datum itself, we generally work with neighborhoods 
of it. Such neighborhoods are typically sets of the form $x + H$ where 
$x$ is some element lying in a $K$-Banach $E$ and $H$ is a lattice in 
$E$. A basic example is the approximate $5$-adic number $1 + O(5^{10})$ 
which corresponds to the neighborhood of $1$ in $\Z_5$, which is the set 
of all $5$-adic numbers congruent to $1$ modulo $5^{10}$. In the context
of computer science, we will often use the notation $x + O(H)$ to refer 
to $x + H$.

@@

\smallskip

@@

\begin{rmk}
In some cases, it would be possible to determine suitable lattices 
$\tilde H_i$ together with their images under $g'_i(x)$ (or, at
least, good approximations of them) before starting the computation 
by using mathematical arguments. If possible, this generally helps a
lot. We shall present in \S \ref{ssec:SOMOS-solution} an example of 
this.
\end{rmk}

@@

@@

Many natural $p$-adic objects do not lie in vector spaces:
points in projective spaces or elliptic curves,
subspaces of a fixed vector space (which lie in Grassmannians),
classes of isomorphism of certain curves (which lie in various moduli spaces),
\emph{etc.}  In this section we extend the formalism 
developed in Section \ref{sec:philosophy} to a more general setting:
we consider the quite general case of strictly
differentiable manifolds locally modeled on ultrametric Banach spaces.
This covers all the 
aforementioned examples.

@@

\subsection{Strictly differentiable $K$-manifolds}
\label{ssec:manifold}

The theory of finite dimensional $K$-manifolds is presented for example 
in \cite{Schneider}, parts 8 and 9. In this Section, we shall work with
a slightly different notion of manifolds which allows also Banach vector 
spaces of infinite dimension.
More precisely, for us, a \emph{strictly differentiable $K$-manifold} 
(or just \emph{$K$-manifold} for short) is the data of a topological 
space $V$ together with an open covering $V = \bigcup_{i \in I} V_i$ 
(where $I$ is some set) and, for all $i \in I$, an homeomorphism 
$\varphi_i : V_i \to U_i$ where $U_i$ is an open subset of a $K$-Banach 
$E_i$ such that for all $i,j \in I$ for which $V_i \cap V_j$ is 
nonempty, the composite map
\begin{equation}
\label{eq:psiij}
\psi_{ij} : 
\varphi_i(V_{ij}) \stackrel{\varphi_i^{-1}}{\longrightarrow} 
V_{ij} \stackrel{\varphi_j}{\longrightarrow} \varphi_j(V_{ij})
\quad \text{(with } V_{ij} = V_i \cap V_j \text{)}
\end{equation}
is strictly differentiable at all points in its domain. We recall that
the mappings $\varphi_i$ above are the so-called \emph{charts} and that
their collection defines an \emph{atlas} of $V$. In the sequel, we shall
assume further that the open covering $V = \bigcup_{i \in I} V_i$ is
locally finite, which means that every point $x \in V$ lies only in a
finite number of $V_i$'s. Trivial examples of $K$-manifolds are 
$K$-Banach themselves.

@@

As usual, we can define the notion of strict differentiability (at some 
point) for a continuous mapping between two $K$-manifolds by viewing it 
through the charts. A strictly differentiable map $f : V \to V'$ induces 
a linear map on tangent spaces $f'(x) : T_x V \to T_{f(x)} V'$ for all 
$x$ in the domain $V$. It is called the \emph{differential} of $f$ at
$x$.

@@

\section{Examples and applications}

In this Section, we illustrate the theory developed in this paper by 
giving examples and applications. In \S \ref{ssec:differentials}, we 
compute the differential of a bunch of usual operations involving 
$p$-adic objects of different nature. Very often, we will observe that 
this computation is easy even if the underlying operation is 
sophisticated (\emph{e.g.} Gr\"obner basis).
In \S \ref{ssec:SOMOS-solution}, we study the SOMOS 4 sequence 
introduced in \S \ref{ssec:stepbystep} and, making a crucial use of 
Lemma \ref{lem:main} and Proposition \ref{prop:locanalytic}, we design
a quite \emph{stable} algorithm for computing it.

\subsection{Differential of usual operations}
\label{ssec:differentials}

As in the previous Sections, we always work over a fixed ultrametric 
field $K$, whose absolute value is denoted by $|\cdot|$. We recall that 
usual examples are the field of $p$-adic numbers $\Q_p$ and the field of 
Laurent series $k((X))$ over a base field $k$.

@@

\subsubsection*{Numbers}

@@

\subsubsection*{Univariate polynomials}

@@

Another usual operations on polynomials is the factorization. Suppose 
that we are given a monic polynomial $P \in K[X]$ written as a product 
$P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. Hensel's 
lemma implies that there exists a small neighborhood $\mathcal U$ of 
$P_0$ such that any monic polynomial $P \in \mathcal U$ factors uniquely 
as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ and $B_0$ 
respectively. Thus, we can consider the application $f : P \mapsto (A,B)$ 
defined on the subset of $\mathcal U$ consisting of monic polynomials.
We want to differentiate $f$ at $P_0$. For this, we differentiate the 
equality $P = A B$ around $P_0$, obtaining 
\begin{equation}
\label{eq:difffactor}
dP = A_0 \cdot dB + B_0 \cdot dA.
\end{equation}
where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).

@@

\subsubsection*{Multivariate polynomials}

We consider the ring $K[\XX] = K[X_1, \ldots, X_n]$ of polynomials in $n$
variables over $K$ and fix a monomial order on it. We then have a 
notion of division in $K[\XX]$: if $f, f_1, \ldots, f_s$ are polynomials
in $K[\XX]$, there exists a writing:
$$f = q_1 f_1 + \cdots + q_s f_s + r$$
where $q_1, \ldots, q_s, r \in K[\XX]$ and no term of $r$ is divisible
by the leading term of some $f_i$ ($1 \leq i \leq s$). Moreover, assuming
that $(f_1, \ldots, f_s)$ is a Gr\"obner basis\footnote{Without this
assumption, there still exists a canonical choice of $r$. We do not know
if the computation of the differential that follows extends to a more
general setting.} (of the ideal generated by these polynomials), the 
polynomial $r$ is uniquely determined and called the \emph{remainder} of 
the division of $f$ by the family $(f_1, \ldots, f_s)$. The application 
$(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and similarly to 
the case of univariate polynomials, we can compute its differential: we 
find that $dr$ is obtained as the remainder of the division of $f - (q_1 
\cdot d f_1 + \cdots + q_s \cdot d f_s)$ by $(f_1, \ldots, f_s)$.

We now focus on the computation of Gr\"obner basis. Recall that any 
ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We can 
then consider the application mapping a family $(f_1, \ldots, f_s)$ of 
\emph{homogeneous}\footnote{This restriction will be convenient for us 
mainly because we are using the article \cite{Vaccon}, which deals only 
with homogeneous polynomials. It is however probably not essential.} 
polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
\ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
application at $(f_1, \ldots, f_s)$ is continuous provided that:
\begin{itemize}
\item the sequence $(f_1, \ldots, f_s)$ is regular
\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
$f_1, \ldots, f_i$ is wearly-$w$ where $w$ denotes the fixed monomial
order (\emph{cf} \cite{Vaccon} for more precisions).
\end{itemize}
A similar argument proves that it is actually differentiable at such
points. We are now going to compute the differential. For this we
remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
g_t)$ generate the same ideal, we have a relation:
\begin{equation}
\label{eq:prodgrobner}
(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
\end{equation}
where $A$ is a $(s \times t)$ matrix with coefficients in $K[\XX]$.
Moreover following Vaccon's construction, we see that the entries of $A$ 
can be chosen in such a way that they define differentiable functions.
Differentiating Eq.~\eqref{eq:prodgrobner}, we get:
$$(d f_1, \ldots, d f_s) \cdot A + (f_1, \ldots, f_s) \cdot dA = 
(d g_1, \ldots, d g_t).$$
Finally, using that we are working with \emph{reduced} Gr\"obner
basis, the above equality implies that $d g_i$ is the remainder in
the division of the $i$-th coefficient of the product
$(d f_1, \ldots, d f_s) \cdot A$ by the family $(g_1, \ldots, g_t)$.

\subsubsection*{Matrices}

@@

We now move to determinants, \emph{i.e.} we compute the differential
of the function $\det : M_n(K) \to K$. It is a standard computation.
To begin with, we consider an invertible matrix $M$ and write:
\begin{eqnarray*}
\det(M + dM) &=& \det(M) \cdot \det(I + M^{-1} \cdot dM)  \\
&=& \det(M) \cdot \big(1 + \tr(M^{-1} \cdot dM)\big) \\
&=& \det(M) + \tr(\com(M) \cdot dM)
\end{eqnarray*}
where $\com(M)$ denotes the comatrix of $M$. The differential of 
$\det$ at $M$ is then $dM \mapsto \tr(\com(M) \cdot dM)$. It turns
out that this formula is still valid when $M$ is not invertible.
The same computation extends readily to characteristic polynomials,
since they are defined as determinants. More precisely, let us 
consider the function $\chi : M_n(K) \to K_{\leq n}[X]$ (where
the latter ring denotes the finite dimensional vector space of 
polynomials over $K$ of degree at most $n$) taking a matrix $M$
to its characteristic polynomial $\det(X-M)$.
Then $\chi$ is differentiable at each point $M \in M_n(K)$ and its 
differential is given by $dM \mapsto \tr(\com(X{-}M) \cdot dM)$.

Other important tools related to matrices are usual factorizations as, 
for example, LU factorization. In what follows, let us agree to define a 
LU factorization of a square matrix $M \in M_n(K)$ as a writing $M = LU$ 
where $L$ is lower triangular and unipotent and $U$ is upper triangular. 
Such a writing exists and is unique provided that all principal minors 
of $M$ do not vanish. We can then consider the mapping $M \mapsto (L,U)$ 
defined over the Zariski-open set of matrices satisfying the above 
condition. In order to differentiate it, we differentiate the relation 
$M = LU$ and rewrite the result as follows:
$$L^{-1} dM \: U^{-1} = L^{-1} \cdot dL + dU \cdot U^{-1}.$$
We remark that in the right hand side of the above formula, the first
summand is lower triangular and all its diagonal entries vanish whereas 
the second summand is upper triangular. Hence in order to compute $dL$
and $dU$, one can proceed as follows: (1)~we compute the product $dX = 
L^{-1} dM \: U^{-1}$, (2)~we separate the lower and upper part of $dX$
obtaining this way $L^{-1} \cdot dL$ and $dU \cdot U^{-1}$ and (3)~we
recover $dL$ and $dU$ by multiplying the above matrices by $L$ on the 
left and $U$ on the right respectively. 
The above discussion extends almost \emph{verbatim} to LUP 
factorization; the only difference is that LUP factorizations are not 
unique but they are on a small neighborhood of $M$ if we fix the matrix 
$P$.

A similar argument works for QR factorization as well.
For us, a QR factorization of a square matrix $M \in M_n(K)$ will be
a writing $M = QR$ where $R$ is unipotent upper triangular and $Q$ is
orthogonal in the sense that $\trans Q \cdot Q$ is diagonal. As before, such 
a writing exists and is unique on a Zariski-open subset of $M_n(K)$. The
mapping $f : M \mapsto (Q,R)$ is then well defined on this subset. We 
would like to emphasize at this point that the orthogonality condition
defines a sub-manifold of $M_n(K)$ which is a apparently \emph{not} a
vector space (it is defined by equations of degree $2$). The codomain
of $f$ is then also a manifold; this example then fits to the setting
of \S \ref{sec:manifold} but not to those of \S \ref{sec:philosophy}. 
The generalization made in \S \ref{sec:manifold} is then needed.
Anyway, we can differentiate $f$ by following the method we used for LU 
factorization: differentiating the relation $M = QR$, we obtain:
\begin{equation}
\label{eq:diffQR}
\trans Q \cdot dM \cdot R^{-1} = \trans Q \cdot dQ + \Delta \cdot dR 
\cdot R^{-1}
\end{equation}
where $\Delta = \trans Q \cdot Q$ is a diagonal matrix by definition.
Moreover by differentiating $\trans Q \cdot Q = \Delta$, we find that
$\trans Q \cdot dQ$ can be written as the sum of an antisymmetric 
matrix and a diagonal one. Since moreove $dR \cdot R^{-1}$ is upper
triangular with all diagonal entries equal to $0$, we see that 
Eq.~\eqref{eq:diffQR} is enough to compute $dQ$ and $dR$ from $Q$, $R$ 
and $dM$.

\subsubsection*{Vector spaces}

@@

Let us now consider the open subset of $M_{n,n-d}(K)$ consisting of 
matrices of full rank. The left kernel defines a mapping $\text{LK} : 
M_{n,n-d}(K) \to \Grass(d,n)$. Let us prove that it is differentiable 
and compute its differential around some point $M \in M_{n,n-d}(K)$. Of 
course, there exists a neighborhood $\mathcal U$ of $M$ whose image is 
entirely contained in a given chart $U_P$. We assume for simplicity that 
$P$ is the identity. On $\mathcal U$, the map $\text{LK}$ corresponds in 
our chart to the application that takes $M$ to the unique matrix $N \in
M_{d, n-d}(K)$ such that $\begin{pmatrix} I_d & N \end{pmatrix} \cdot M 
= 0$. The implicit function theorem then implies that $\text{LK}$ is
differentiable. Furthermore, its differential satisfies the relation
$\begin{pmatrix} 0 & dN \end{pmatrix} \cdot M +
\begin{pmatrix} I_d & N \end{pmatrix} \cdot dM = 0$,
from which we can compute $dN$ by projecting on the $(n-d)$ last
columns.

>>>>

Lemma \ref{lem:main} and Proposition \ref{prop:locanalytic} have 
important applications to effective computations with $p$-adic numbers 
or power series, that we are going to discuss now. Let us recall first 
that a datum involving elements of a complete ultrametric field $K$ 
(\emph{e.g.} $p$-adic numbers or power series) cannot be stored entirely 
on a computed because they basically convey an infinite amount of 
informations. That is why, instead of this datum itself, we generally 
work with neighborhoods of it. In what follows, we shall consider 
neighborhoods of the form $x + H$ where $x$ is some element lying in a 
$K$-Banach $E$ and $H$ is a lattice in $E$. A very basic example is the 
approximate $p$-adic number $1 + O(p^N)$ which corresponds to the 
neighborhood of $1$ in $\Z_p$, which is the set of all $p$-adic numbers 
congruent to $1$ modulo $p^N$. We nevertheless emphasize that our 
definition is quite flexible and allows many neighborhoods that we do 
not consider in general. For instance, an ``admissible'' neighborhood 
around a monic polynomial $P_0 \in \Q_p[X]$ of degree $d$ may be defined 
by congruences of the form $P(a_i) \equiv P_0(a_i) \pmod {p^{n_i}}$ ($1 
\leq i \leq d$) for some fixed elements $a_i \in \Q_p$ and relative 
integers $n_i$.

In the context of computer science, we will often use the notation $x + 
O(H)$ to refer to $x + H$.

@@

@@

\begin{rmk}
In some cases, it is actually possible to determine suitable lattices 
$\tilde H_i$ together with their images under $g'_i(x)$ (or, at
least, good approximations of them) before starting the computation 
by using mathematical arguments. If possible, this generally helps a
lot. We shall present in \S \ref{ssec:SOMOS-solution} an example of 
this.
\end{rmk}

@@

\subsection{Application to SOMOS sequence}
\label{ssec:SOMOS-solution}

We illustrate the theory developed in this paper by giving a simple 
toy application. Other applications will be discussed in subsequent 
articles. More precisely, we study the SOMOS 4 sequence introduced in \S 
\ref{ssec:stepbystep}. Making a crucial use of Lemma \ref{lem:main} and 
Proposition \ref{prop:locanalytic}, we design a \emph{stable} algorithm 
for computing it.

From now, we will always consider SOMOS sequences with values in $\Q_p$ 
(for some prime number $p$) and assume that $u_0, u_1, u_2, u_3$ are 
chosen such that all terms $u_n$'s are determined (\emph{i.e.} there is 
no division by zero). We assume also for simplicity that $u_0, u_1, u_2, 
u_3$ are all units in $\Z_p$. By Laurent phenomenon, this implies first 
that all $u_n$'s lie in $\Z_p$ and second that if $u_0, u_1, u_2, u_3$ 
are known with finite precision $O(p^N)$ then all terms $u_n$'s are 
known with the same precision. Algorithm \ref{algo:SOMOS} presented on 
page \pageref{algo:SOMOS} performs this computation.

\begin{algorithm}[t]
\SetKwInOut{Assumption}{Assumption}
\KwIn{$a,b,c,d$ --- four initial terms of a SOMOS 4 sequence $(u_n)_{n \geq 0}$}
\KwIn{$n, N$ --- two integers}
\Assumption{$a$, $b$, $c$ and $d$ lie in $\Z_p^\times$ and are known at 
precision $O(p^N)$}
\Assumption{None of $u_i$ $(0 \leq i \leq n$) is divisible by $p^N$}
\KwOut{$u_n$ at precision $O(p^N)$}
\BlankLine
$\prec$ $\leftarrow$ $N$\;
\For {$i$ from $1$ to $n-3$}{
  $\prec$ $\leftarrow$ $\prec + v_p(bd + c^2) - v_p(a)$\;
  {\bf lift} (arbitrary) 
    $b$, $c$ and $d$ to precision $O(p^{\prec})$\label{line:lift}\;
  $\prec$ $\leftarrow$ $\prec - v_p(a)$\;
  $e$ $\leftarrow$ $\frac{bd+c^2} a$; \hspace{1cm}
   \tcp{$e$ is known at precision $O(p^\prec)$}
  $a, b, c, d$ $\leftarrow$ $b + O(p^{\prec}), c + O(p^{\prec}), d + O(p^{\prec}), e$\;
}
\Return{$d + O(p^N)$}\;
\caption{\sc SOMOS$(a, b, c, d, n, N)$}\label{algo:SOMOS}
\end{algorithm}

We now prove that it is correct.
We introduce the (partial) function $f : \Q_p^4 \to \Q_p^4$ defined by 
$f(a,b,c,d) = (b,c,d,\frac{bd+c^2}a)$, so that, for all $i$, we have 
$(u_i, u_{i+1}, u_{i+2}, u_{i+3}) = f_i(u_0, u_1, u_2, u_3)$ where $f_i
= f \circ \cdots \circ f$ ($i$ times). Clearly, $f$ is  
differentiable on $\Q_p^\times \times \Q_p^3$ and its differential in the 
canonical basis is given by the matrix:
$$D(a,b,c,d) = \begin{pmatrix}
0 & 1 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 \\
-\frac{bd+c^2}{a^2} & \frac d a & \frac {2c} a & \frac b a
\end{pmatrix}$$
whose determinant is $\frac{bd+c^2}{a^2}$. Thus, if the $(i+4)$-th term 
of the SOMOS sequence is defined, the mapping $f_i$ is differentiable
at $(u_0, u_1, u_2, u_3)$ and its differential $\varphi_i = 
f'_i(u_0, u_1, u_2, u_3)$ is given by the matrix:
$$D_i = D(u_{i-1}, u_i, u_{i+1}, u_{i+2}) \cdots
D(u_1, u_2, u_3, u_4) \cdot D(u_0, u_1, u_2, u_3).$$
Moreover thanks to the Laurent phenomenon, we know that $D_i$ has
integral coefficients, \emph{i.e.} $\varphi_i$ stabilizes the lattice
$\Z_p^4$.
We are now going to prove by induction on $i$ that, at the end of the 
$i$-th iteration of the loop, we have $\prec = N + v_p(\det D_i)$ and
\begin{equation}
\label{eq:congrSOMOS}
(a, b, c, d) \equiv (u_i, u_{i+1}, u_{i+2}, u_{i+3}) \pmod
{p^N \varphi_i(\Z_p^4)}.
\end{equation}
The first point is easy. Indeed, from $D_i = D(u_{i-1}, u_i, u_{i+1}, 
u_{i+2}) D_{i-1}$, we deduce $\det D_i = \det D_{i-1} \cdot 
\frac{u_i}{u_{i-3}}$ and the assertion follows by taking determinants
and using the induction hypothesis. 
Let us now establish Eq.~\eqref{eq:congrSOMOS}. To avoid confusion, 
let us agree to denote $a'$, $b'$, $c'$, $d'$ and $\prec'$ the values of 
$a$, $b$, $c$, $d$ and $\prec$ respectively at the \emph{beginning} of 
the $i$-th iteration of the loop. By induction hypothesis (or by 
initialization if $i = 1$), we have:
\begin{equation}
\label{eq:congrSOMOS2}
(a', b', c', d') \equiv (u_{i-1}, u_i, u_{i+1}, u_{i+2}) \pmod
{p^N \varphi_{i-1}(\Z_p^4)}.
\end{equation}
Moreover, we know that the determinant of $\varphi_{i-1}$ has valuation
$\prec'$. Hence \eqref{eq:congrSOMOS2} remains true if $a'$, $b'$, $c'$
and $d'$ are replaced by other values which are congruent to them modulo
$p^{\prec'}$. In particular it holds if $a'$, $b'$, $c'$ and $d'$ denotes
the values of $a$, $b$, $c$ and $d$ after the execution of line
\ref{line:lift}. Now applying Lemma \ref{lem:main} and Proposition
\ref{prop:locanalytic} to $\varphi_{i-1}$ and $\varphi_i$ (at the point 
$(u_0, u_1, u_2, u_3)$), we get:
$$f\big((u_{i-1}, u_i, u_{i+1}, u_{i+2}) + p^N \varphi_{i-1}(\Z_p^4)\big)
= (u_i, u_{i+1}, u_{i+2}, u_{i+3}) + p^N \varphi_i(\Z_p^4).$$
By the discussion above, this implies in particular that $f(a',b',c',d')$
belongs to $(u_i, u_{i+1}, u_{i+2}, u_{i+3}) + p^N \varphi_i(\Z_p^4)$. We
conclude by remarking first that $(a,b,c,d) \equiv f(a',b',c',d') \pmod 
{p^\prec \Z_p^4}$ by construction and second that $p^\prec \Z_p^4 \subset
p^N \varphi_i(\Z_p^4)$.

\appendix

@@

Many natural $p$-adic objects do not lie in vector spaces:
points in projective spaces or elliptic curves,
subspaces of a fixed vector space (which lie in Grassmannians),
classes of isomorphism of certain curves (which lie in various moduli spaces),
\emph{etc.}  In this section we extend the formalism 
developed in Section \ref{sec:philosophy} to a more general setting:
we consider the quite general case of 
differentiable manifolds locally modeled on ultrametric Banach spaces.
This covers all the 
aforementioned examples.

@@

\subsection{Differentiable $K$-manifolds}
\label{ssec:manifold}

The theory of finite dimensional $K$-manifolds is presented for example 
in \cite{Schneider}, parts 8 and 9. In this Section, we shall work with
a slightly different notion of manifolds which allows also Banach vector 
spaces of infinite dimension.
More precisely, for us, a \emph{ differentiable $K$-manifold} 
(or just \emph{$K$-manifold} for short) is the data of a topological 
space $V$ together with an open covering $V = \bigcup_{i \in I} V_i$ 
(where $I$ is some set) and, for all $i \in I$, an homeomorphism 
$\varphi_i : V_i \to U_i$ where $U_i$ is an open subset of a $K$-Banach 
$E_i$ such that for all $i,j \in I$ for which $V_i \cap V_j$ is 
nonempty, the composite map
\begin{equation}
\label{eq:psiij}
\psi_{ij} : 
\varphi_i(V_{ij}) \stackrel{\varphi_i^{-1}}{\longrightarrow} 
V_{ij} \stackrel{\varphi_j}{\longrightarrow} \varphi_j(V_{ij})
\quad \text{(with } V_{ij} = V_i \cap V_j \text{)}
\end{equation}
is  differentiable at all points in its domain. We recall that
the mappings $\varphi_i$ above are the so-called \emph{charts} and that
their collection defines an \emph{atlas} of $V$. In the sequel, we shall
assume further that the open covering $V = \bigcup_{i \in I} V_i$ is
locally finite, which means that every point $x \in V$ lies only in a
finite number of $V_i$'s. Trivial examples of $K$-manifolds are 
$K$-Banach themselves.

@@

As usual, we can define the notion of strict differentiability (at some 
point) for a continuous mapping between two $K$-manifolds by viewing it 
through the charts. A  differentiable map $f : V \to V'$ induces 
a linear map on tangent spaces $f'(x) : T_x V \to T_{f(x)} V'$ for all 
$x$ in the domain $V$. It is called the \emph{differential} of $f$ at
$x$.

@@

\section{Differential of usual operations}
\label{ssec:differentials}

We continue to work over a fixed ultrametric field $K$, whose absolute 
value is denoted by $|\cdot|$. We recall that usual examples of such $K$ 
are the field of $p$-adic numbers $\Q_p$ and the field of Laurent series 
$k((X))$ over a base field $k$. The aim of this appendix is to compute 
the differential of many usual operations on numbers, polynomials and 
matrices. Surprisingly, we observe that all differentials we will 
consider are rather easy to compute even if the underlying operation is 
quite involved (\emph{e.g.} Gr\"obner basis).

@@

\subsection{Numbers}

@@

\subsection{Univariate polynomials}

For any integer $d$, let us denote by $K_{< d}[X]$ the set of 
polynomials over $K$ of degree $< d$. It is a finite dimensional vector 
space of dimension $d+1$. The affine space $X^d + K_{< d}[X]$ is then 
the set of monic polynomials over $K$ of degree $d$. We denote it by 
$K_d[X]$.

\subsubsection*{Euclidean division}

@@

\subsubsection*{Subresultants and B\'ezout coefficients}

\todo{To be written.}

\subsubsection*{Factorization}

Suppose that we are given a monic polynomial $P \in K[X]$ written as a 
product $P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. 
Hensel's lemma implies that there exists a small neighborhood $\mathcal 
U$ of $P_0$ such that any monic polynomial $P \in \mathcal U$ factors 
uniquely as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ 
and $B_0$ respectively. Thus, we can consider the application $f : P 
\mapsto (A,B)$ defined on the subset of $\mathcal U$ consisting of monic 
polynomials. We want to differentiate $f$ at $P_0$. For this, we 
differentiate the equality $P = A B$ around $P_0$, obtaining
\begin{equation}
\label{eq:difffactor}
dP = A_0 \cdot dB + B_0 \cdot dA.
\end{equation}
where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).

@@

\subsection{Multivariate polynomials}

We consider the ring $K[\XX] = K[X_1, \ldots, X_n]$ of polynomials in $n$
variables over $K$ and fix a monomial order on it.

\subsubsection*{Division}

We then have a notion of division in $K[\XX]$: if $f, f_1, \ldots, f_s$ 
are polynomials in $K[\XX]$, there exists a writing:
$$f = q_1 f_1 + \cdots + q_s f_s + r$$
where $q_1, \ldots, q_s, r \in K[\XX]$ and no term of $r$ is divisible
by the leading term of some $f_i$ ($1 \leq i \leq s$). Moreover, assuming
that $(f_1, \ldots, f_s)$ is a Gr\"obner basis\footnote{Without this
assumption, there still exists a canonical choice of $r$. We do not know
if the computation of the differential that follows extends to a more
general setting.} (of the ideal generated by these polynomials), the 
polynomial $r$ is uniquely determined and called the \emph{remainder} of 
the division of $f$ by the family $(f_1, \ldots, f_s)$. The application 
$(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and similarly to 
the case of univariate polynomials, we can compute its differential: we 
find that $dr$ is obtained as the remainder of the division of $f - (q_1 
\cdot d f_1 + \cdots + q_s \cdot d f_s)$ by $(f_1, \ldots, f_s)$.

\subsubsection*{Gr\"obner basis}

Recall that any 
ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We can 
then consider the application mapping a family $(f_1, \ldots, f_s)$ of 
\emph{homogeneous}\footnote{This restriction will be convenient for us 
mainly because we are using the article \cite{Vaccon}, which deals only 
with homogeneous polynomials. It is however probably not essential.} 
polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
\ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
application at $(f_1, \ldots, f_s)$ is continuous provided that:
\begin{itemize}
\item the sequence $(f_1, \ldots, f_s)$ is regular
\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
$f_1, \ldots, f_i$ is wearly-$w$ where $w$ denotes the fixed monomial
order (\emph{cf} \cite{Vaccon} for more precisions).
\end{itemize}
A similar argument proves that it is actually differentiable at such
points. We are now going to compute the differential. For this we
remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
g_t)$ generate the same ideal, we have a relation:
\begin{equation}
\label{eq:prodgrobner}
(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
\end{equation}
where $A$ is a $(s \times t)$ matrix with coefficients in $K[\XX]$.
Moreover following Vaccon's construction, we see that the entries of $A$ 
can be chosen in such a way that they define differentiable functions.
Differentiating Eq.~\eqref{eq:prodgrobner}, we get:
$$(d g_1, \ldots, d g_t) =
(d f_1, \ldots, d f_s) \cdot A + (f_1, \ldots, f_s) \cdot dA.$$
Finally, using that we are working with \emph{reduced} Gr\"obner
basis, the above equality implies that $d g_i$ is the remainder in
the division of the $i$-th coefficient of the product
$(d f_1, \ldots, d f_s) \cdot A$ by the family $(g_1, \ldots, g_t)$.

\subsection{Matrices}

@@

\subsubsection*{Determinants and characteristic polynomials}

Let us first compute the differential of the function $\det : M_n(K) \to 
K$. It is a standard computation. To begin with, we consider an 
invertible matrix $M$ and write:
\begin{eqnarray*}
\det(M + dM) &=& \det(M) \cdot \det(I + M^{-1} \cdot dM)  \\
&=& \det(M) \cdot \big(1 + \tr(M^{-1} \cdot dM)\big) \\
&=& \det(M) + \tr(\com(M) \cdot dM)
\end{eqnarray*}
where $\com(M)$ denotes the comatrix of $M$. The differential of 
$\det$ at $M$ is then $dM \mapsto \tr(\com(M) \cdot dM)$. It turns
out that this formula is still valid when $M$ is not invertible.
The same computation extends readily to characteristic polynomials,
since they are defined as determinants. More precisely, let us 
consider the function $\chi : M_n(K) \to K_{\leq n}[X]$ (where
the latter ring denotes the finite dimensional vector space of 
polynomials over $K$ of degree at most $n$) taking a matrix $M$
to its characteristic polynomial $\det(X-M)$.
Then $\chi$ is differentiable at each point $M \in M_n(K)$ and its 
differential is given by $dM \mapsto \tr(\com(X{-}M) \cdot dM)$.

Let us agree to define a 
LU factorization of a square matrix $M \in M_n(K)$ as a writing $M = LU$ 
where $L$ is lower triangular and unipotent and $U$ is upper triangular. 
Such a writing exists and is unique provided that all principal minors 
of $M$ do not vanish. We can then consider the mapping $M \mapsto (L,U)$ 
defined over the Zariski-open set of matrices satisfying the above 
condition. In order to differentiate it, we differentiate the relation 
$M = LU$ and rewrite the result as follows:
$$L^{-1} dM \: U^{-1} = L^{-1} \cdot dL + dU \cdot U^{-1}.$$
We remark that in the right hand side of the above formula, the first
summand is lower triangular and all its diagonal entries vanish whereas 
the second summand is upper triangular. Hence in order to compute $dL$
and $dU$, one can proceed as follows: (1)~we compute the product $dX = 
L^{-1} dM \: U^{-1}$, (2)~we separate the lower and upper part of $dX$
obtaining this way $L^{-1} \cdot dL$ and $dU \cdot U^{-1}$ and (3)~we
recover $dL$ and $dU$ by multiplying the above matrices by $L$ on the 
left and $U$ on the right respectively. 
The above discussion extends almost \emph{verbatim} to LUP 
factorization; the only difference is that LUP factorizations are not 
unique but they are on a small neighborhood of $M$ if we fix the matrix 
$P$.

\subsubsection*{QR factorization}

For us, a QR factorization of a square matrix $M \in M_n(K)$ will be
a writing $M = QR$ where $R$ is unipotent upper triangular and $Q$ is
orthogonal in the sense that $\trans Q \cdot Q$ is diagonal. As before, such 
a writing exists and is unique on a Zariski-open subset of $M_n(K)$. The
mapping $f : M \mapsto (Q,R)$ is then well defined on this subset. We 
would like to emphasize at this point that the orthogonality condition
defines a sub-manifold of $M_n(K)$ which is a apparently \emph{not} a
vector space (it is defined by equations of degree $2$). The codomain
of $f$ is then also a manifold; this example then fits to the setting
of \S \ref{sec:manifold} but not to those of \S \ref{sec:philosophy}. 
The generalization made in \S \ref{sec:manifold} is then needed.
Anyway, we can differentiate $f$ by following the method we used for LU 
factorization: differentiating the relation $M = QR$, we obtain:
\begin{equation}
\label{eq:diffQR}
\trans Q \cdot dM \cdot R^{-1} = \trans Q \cdot dQ + \Delta \cdot dR 
\cdot R^{-1}
\end{equation}
where $\Delta = \trans Q \cdot Q$ is a diagonal matrix by definition.
Moreover by differentiating $\trans Q \cdot Q = \Delta$, we find that
$\trans Q \cdot dQ$ can be written as the sum of an antisymmetric 
matrix and a diagonal one. Since moreove $dR \cdot R^{-1}$ is upper
triangular with all diagonal entries equal to $0$, we see that 
Eq.~\eqref{eq:diffQR} is enough to compute $dQ$ and $dR$ from $Q$, $R$ 
and $dM$.

\subsection{Vector spaces}

@@

\subsubsection*{Left kernels}
